An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-11-17 paper-draft/paper.tex
udpate bibliography
add chapter on near-toeplitz solver

Thus, in the best case ($n$ parallel processors),
cyclic reduction requires 
$2log(n) - 1$ steps.
For even moderately large $n$,
this is significantly smaller than
the $2n$ steps required by the classical Thomas algorithm.
This makes cyclic reduction a good fit
for massively parallel architectures like GPUs.

@@

When the matrix $A$ is near-Toeplitz tridiagonal,
if we substitute in these equations:

>>>>

Thus, in the best case ($n$ parallel processors),
cyclic reduction requires 
$2log_2(n) - 1$ steps.
For even moderately large $n$,
this is significantly smaller than
the $2n$ steps required by the classical Thomas algorithm.
This makes cyclic reduction a good fit
for massively parallel architectures like GPUs.

@@

When the matrix $A$ is near-Toeplitz tridiagonal,
if we substitute in the Eqs.
Eqs. (\ref{eqn:forward-reduction-1}) -
(\ref{eqn:forward-reduction-3}):

### 2015-11-17 thesis/introduction-and-background.tex
add chapter on near-toeplitz solver

\begin{align}
    f^{\prime}_1 + 2f^{\prime}_2 = \frac{-5f_1 + 4f_2 + f_3}{dx} \\
    f^{\prime}_{n} + 2f^{\prime}_{n-1}
    =
    \frac{5f_{n} - 4f_{n-2} -  f_{n-1}}{dx}
\end{align}
%
The resulting tridiagonal system is then:

@@

As mentioned in \ref{sec:tridiagonal-compact-schemes},
this work is concernted with
\emph{tridiagonal} compact finite differences
for their relative ease in evaluation.
The resulting tridiagonal system must be solved
repeatedly for all lines of grid points in the computational grid.
For a regular Cartesian grid (generally employed in DNS),
the coefficients of the tridiagonal systems
are the same for each grid line,
and only the right hand sides are different.
Thus, the compact finite difference evaluation
effectively requires the solution of a tridiagonal system
for several right hand sides.
In this section,
we describe the applicability of several algorithms
for solving this problem on the GPU.

@@

\begin{align} 
& a^{\prime}_i = -a_{i-1}k_1 \
    \label{eqn:forward-reduction-1}& \\
& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 \
    \label{eqn:forward-reduction-2}& \\
& c^{\prime}_i = -c_{i+1}k_2 \
    \label{eqn:forward-reduction-3}& \\
& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 \
    \label{eqn:forward-reduction-4}&
\end{align}
%
where,

\begin{align}
& k_1 = \frac{a_i}{b_{i-1}} \label{eqn:k1-update}& \\
& k_2 = \frac{c_i}{b_{i+1}} \label{eqn:k2-update}&
\end{align}

@@

\begin{align} \label{eqn:backward-substitution}
x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - \
    c^{\prime}_ix_{i+1}}{b^{\prime}_i}
\end{align}
%
For the last index $i=n$,
the forward reduction step is instead:
\begin{align} \label{eqn:forward-reduction-last}
    & a^{\prime}_n = -a_{n-1}k_1 & \\
    & b^{\prime}_n = b_n - c_{n-1}k_1 & \\
    & d^{\prime}_n = d_n - d_{n-1}k_1&
\end{align}
%
And for $i=1$, the backward substitution step is instead:

\begin{align} \label{eqn:backward-substitution-first}
x_1 = \frac{d^{\prime}_1 - c^{\prime}_1x_{2}}{b^{\prime}_1}
\end{align}

@@

\subsection{Cyclic reduction implementation on GPUs}

@@

the threads assigned to each even index $i$
compute the coefficients
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$.
In practice, the coefficients are updated \emph{in-place}.
Figure \ref{fig:forward-reduction-step} shows the updates
to the coefficient array $\bm{b}$
in the first forward reduction step).
In each step of forward reduction,
a thread accesses values from the coefficient arrays
$\bm{a}$, $\bm{b}$ and $\bm{c}$ in a \emph{strided} fashion.
At every subsequent step,
this stride is doubled, while the number of active threads is halved
\ref{fig:cyclic-reduction}.
In the backward substitution phase,
the strides are halved at each step,
while the number of active threads is doubled.

This \emph{hourglass} pattern results in low GPU utilization
towards the end of forward reduction
and the beginning of back substitution.
Further,
the long-strided memory accesses towards the end
of the forward reduction
and the beginning of backward substitution
is problematic for both
global memory and shared memory implementations.
The global memory access is \emph{uncoalesced},
leading to severe performance penalties.
The use of shared memory can prevent
long-strided global memory access.
Unfortunately,
the power-of-two strides at each successive
leads to bank conflicts,
as described in Sec. \ref{subsubsec:shared-memory}.
The bank conflicts become increasingly severe
at each forward reduction phase,
and decreasingly so during the back substitution phase.
The limited amount of shared memory places restricions
on the size of tridiagonal systems that can be solved,
and also on the number of blocks that can run in parallel.

Despite these problems,
cyclic reduction remains an attractive algorithm
for GPUs,
for its low algorithmic complexity,
and lower work per step compared to PCR.
Much work has been done on
optimizing cyclic reduction performance on GPUs,
and addressing the issues discussed.
Zhang et al. \cite{Zhang2010FTS} propose a
\emph{hybrid} solver
that uses both cyclic reduction and parallel cyclic reduction
to reduce bank conflicts.
G{\"o}dekke et al. \cite{GoSt11CR}
use a method of separately storing
the even and odd indexed equations
to arrive at a bank-conflict free solver
at the cost of additional shared memory usage.
Davidson et al. \cite{davidson2011register}
describe the method of
\emph{register packing}---performing more computations
on \emph{registers}, rather than shared memory---as
a means to reduce shared memory usage in cyclic reduction.
Esfahanian et. al, \cite{esfahanian2014efficient}
avoid shared memory (and associated bank conflicts entirely)
using a data rearrangement scheme to improve global memory access.
Our approach takes a different route,
and is focussed on near-Toeplitz systems---exploiting
the relatively simple matrix structure
to reduce the number of computations and memory accesses
performed by each thread
at every cyclic reduction step.

>>>>

\begin{align}
    f^{\prime}_1 + 2f^{\prime}_2 &= \frac{-5f_1 + 4f_2 + f_3}{dx} \\
    f^{\prime}_{n} + 2f^{\prime}_{n-1}
    &=
    \frac{5f_{n} - 4f_{n-2} -  f_{n-1}}{dx}
\end{align}
%
The resulting tridiagonal system is then:

@@

As mentioned in \ref{sec:tridiagonal-compact-schemes},
this work is concerned with
\emph{tridiagonal} compact finite differences
for their relative ease in evaluation.
The resulting tridiagonal system must be solved
repeatedly for all lines of grid points in the computational grid.
For a regular Cartesian grid (generally employed in DNS),
the coefficients of the tridiagonal systems
are the same for each grid line,
and only the right hand sides are different.
Thus, the compact finite difference evaluation
effectively requires the solution of a tridiagonal system
for several right hand sides.
In this section,
we describe the applicability of some algorithms
for solving this problem on the GPU.
We refer to ~\cite{chang2014guide} for
a more comprehensive list of algorithms,
along with details of their GPU implementation.

@@

\begin{align} 
    a^{\prime}_i &= -a_{i-1}k_1 \
    \label{eqn:forward-reduction-1} \\
    b^{\prime}_i &= b_i - c_{i-1}k_1 - a_{i+1}k_2 \
    \label{eqn:forward-reduction-2} \\
    c^{\prime}_i &= -c_{i+1}k_2 \
    \label{eqn:forward-reduction-3} \\
    d^{\prime}_i &= d_i - d_{i-1}k_1  - d_{i+1}k_2 \
    \label{eqn:forward-reduction-4}
\end{align}
%
where,

\begin{align}
    k_1 &= \frac{a_i}{b_{i-1}} \label{eqn:k1-update} \\
    k_2 &= \frac{c_i}{b_{i+1}} \label{eqn:k2-update}
\end{align}

@@

\begin{equation} \label{eqn:backward-substitution}
x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - \
    c^{\prime}_ix_{i+1}}{b^{\prime}_i}
\end{equation}
%
For the last index $i=n$,
the forward reduction step is instead:
\begin{align} \label{eqn:forward-reduction-last}
    a^{\prime}_n &= -a_{n-1}k_1  \\
    b^{\prime}_n &= b_n - c_{n-1}k_1  \\
    d^{\prime}_n &= d_n - d_{n-1}k_1
\end{align}
%
And for $i=1$, the backward substitution step is instead:

\begin{equation} \label{eqn:backward-substitution-first}
x_1 = \frac{d^{\prime}_1 - c^{\prime}_1x_{2}}{b^{\prime}_1}
\end{equation}
%
In practice, the right-hand side vector can be safely
overwritten with the solution values
in backward substitution.

@@

\subsection{Cyclic reduction implementation on GPUs}
\label{subsec:cyclic-reduction-gpu-implementation}

@@

the threads assigned to each even index $i$
compute the coefficients
and right hand side for the reduced tridiagonal system
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
and $d_i^\prime$.
In practice, the coefficients and right hand side
are updated \emph{in-place}.
Figure \ref{fig:forward-reduction-step} shows the updates
to the coefficient array $\bm{b}$
in the first forward reduction step,
a similar pattern is seen for the arrays
$\bm{a}$, $\bm{c}$ and $\bm{d}$.
In each step of forward reduction,
a thread accesses values from the
coefficient arrays and right hand side
in a \emph{strided} fashion.
At every subsequent step,
this stride is doubled, while the number of active threads is halved
\ref{fig:cyclic-reduction}.
In the backward substitution phase,
the strides are halved at each step,
while the number of active threads is doubled.

Thus, the following issues are encountered
while implementing cyclic reduction for GPUs:

\begin{enumerate}
    \item GPU utilization is low towards the end of forward reduction,
        and in the beginning of backward substitution.

    \item Because coefficients and right hand sides
        are updated \emph{in-place},
        synchronization between the blocks is required at
        the end of each step.

    \item If the threads work entirely with global memory,
        memory accesses are increasingly
        \emph{uncoalesced} in the forward reduction phase
        (and become increasingly coalesced in the
        backward substitution phase).

    \item The use of shared memory prevents
        uncoalesced global memory access.
        Unfortunately,
        the power-of-two strides at each successive
        leads to bank conflicts,
        as described in Sec. \ref{subsubsec:shared-memory}.
        The bank conflicts become increasingly severe
        at each forward reduction phase,
        and decreasingly so during the back substitution phase.

    \item The limited amount of shared memory places restricions
        on the size of tridiagonal systems that can be solved,
        and also on the number of systems that can be solved
        concurrently.
\end{enumerate}

Despite these issues,
cyclic reduction remains an attractive algorithm
for GPUs,
for its low algorithmic complexity,
and lower work per step compared to PCR.
Much work has been done on
optimizing cyclic reduction performance on GPUs,
and addressing the issues discussed.
Zhang et al. \cite{Zhang2010FTS} propose a
\emph{hybrid} solver
that uses both cyclic reduction and parallel cyclic reduction
to reduce the number and severity of bank conflicts,
and also to have better thread activity overall.
G{\"o}dekke et al. \cite{GoSt11CR}
use a method of separately storing
the even and odd indexed equations
to arrive at a bank-conflict free solver
at the cost of additional shared memory usage.
Davidson et al. \cite{davidson2011register}
describe the method of
\emph{register packing}---performing more computations
on \emph{registers}, rather than shared memory---as
a means to reduce shared memory usage in cyclic reduction.
Esfahanian et. al, \cite{esfahanian2014efficient}
avoid shared memory (and associated bank conflicts entirely)
using a data rearrangement scheme to improve global memory access.
Our approach takes a different route,
and is focussed on \emph{near-Toeplitz} systems---exploiting
the relatively simple matrix structure
to reduce the number of computations and memory accesses
performed by each thread
at every cyclic reduction step.

### 2015-11-17 thesis/proposed-tridiagonal-algorithm.tex
add chapter on near-toeplitz solver

\section{Modified cyclic reduction}
\subsection{Analysis of forward reduction coefficiets}
\subsection{Precomputing coefficients}
\subsection{Forward reduction}
\subsection{Back substitution}

\section{Implementation}
\subsection{General strategy}
\subsection{Global memory implementation}
\subsubsection{Algorithm}
\subsubsection{Kernels}
\subsubsection{Grid and thread block configuration}

\subsection{Shared memory implementation}
\subsubsection{Algorithm}
\subsubsection{Kernel}
\subsubsection{Grid and thread block configuration}

>>>>

Here, we explore the idea of
exploiting the relatively simple matrix structure
of the coefficient matrix appearing in
tridiagonal compact finite difference schemes.
The coefficient matrix $A$ for tridiagonal schemes
has the following general form:

\begin{equation} \label{eqn:toeplitz-matrix}
A = 
\begin{pmatrix}
     b_1 & c_1  \\
     a_0 & b_0  &  c_0  \\
         & a_0  &  b_0 &  c_0  \\
         &      &  a_0 &  b_0 &  c    \\
         &      &      &      &  \ddots \\
         &      &      &      &     &  \ddots  \\
         &      &      &      &     &  a_n  &  b_n
\end{pmatrix}
\end{equation}
%
where, in general
\begin{align*}
    a_0 &\neq a_n  \\
    b_1 &\neq b_0 \neq b_n \\
    c_1 &\neq c_0 
\end{align*}
%
We refer to matrices with this specific structure as
\emph{near-Toeplitz tridiagonal matrices},
and the corresponding linear systems as
\emph{near-Toeplitz tridiagonal systems}.
These matrices appear in a wide range of applications
~\cite{sun1995application}
such as alternating direction implicit methods,
line relaxation methods,
and numerical solutions to one-dimensional differential equations.
In this chapter,
we will describe an efficient solver for solving such
near-Toeplitz tridiagonal systems efficiently on GPUs.

\section{Modified cyclic reduction for near-Toeplitz systems}

\subsection{Forward reduction}

The forward reduction phase reduces
a $n-by-n$ system of equations to a
$2-by-2$ system of equations in $log_2(n)-1$ steps,
by applying
Eqs. (\ref{eqn:forward-reduction-1}) -
(\ref{eqn:forward-reduction-4})
to every even-indexed equation at each step.
When solving the tridiagonal systems
with the same coefficient matrix,
but repeatedly for different right hand sides,
we note that the results of
Eqs. (\ref{eqn:forward-reduction-1}) -
(\ref{eqn:forward-reduction-3})
remain unchanged for the different right hand sides.
These results correspond to the
coefficients of the tridiagonal system
produced at each forward reduction step.
Thus, given a tridiagonal system,
we may precompute
the coefficients of all
the reduced systems appearing in the
forward reduction steps,
and reuse them for each right-hand side.
We note that for a general tridiagonal system,
this requires storage for $3.(2N - 2)$ coefficients,
in addition to the $3N$ coefficients
for the original tridiagonal system.

Let us consider the case
when the tridiagonal system is \emph{near-Toeplitz},
i.e., when the coefficient matrix is of the form
$A$ in Eq. \ref{eqn:toeplitz-matrix}.
We examine the effect of the
first forward reduction step by making the following substitutions
in Eqs. (\ref{eqn:forward-reduction-1}) -
(\ref{eqn:forward-reduction-3}):

\begin{align*}
    a_2 &= a_3 = a_4 = \hdots &\equiv a_0 \\
    b_2 &= b_3 = b_4 = \hdots &\equiv b_0 \\
    c_2 &= c_3 = c_4 = \hdots &\equiv c_0
\end{align*}
%
We observe that the resulting coefficients
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
correspond to the coefficients of a tridiagonal matrix with
exactly the near-Toeplitz structure of $A$.
This form-preserving property of cyclic reduction
has been reported for block Toeplitz tridiagonal systems
by Bini et al. \cite{bini}.

The fact that the reduced system at each step
is near-Toeplitz can be exploited
to reduce the cost of
precomputing and storing the coefficients.
Each near-Toeplitz matrix is completely defined
by only a handful of coefficients:
$\{b_1, c_1, a_0, b_0, c_0, a_n, b_n\}$,
making its storage extremely compact
compared to the case of general tridiagonal systems.
In addition,
it is advantageous to precompute and store the
auxiliary variables $k_1$ and $k_2$,
which can similarly be stored compactly.
With all the forward reduction coefficient matrices
and auxillary variables precomputed and stored,
the $m$th forward reduction step for equation $i$
is reduced only to the right hand side update:

\begin{equation}
d^{\prime}_i = d_i - d_{i-1}k_1^{m}  - d_{i+1}k_2^{m}
\label{eqn:precomputed-forward-reduction-step}
\end{equation}
%
where $k_1^m$ and $k_2^m$ are
precomputed values of $k_1$ and $k_2$
for all ``inner'' equations at the $m$th step.
For the ``outer'' equations $i=2$ and $i=n$,
we have instead:
%
\begin{align}
    d^{\prime}_2 &= d_2 - d_{1}k_{1,1}^{m}  - d_{3}k_2^{m} \
    \label{eqn:precomputed-forward-reduction-step-2} \\
    d^{\prime}_n &= d_n - d_{n-1}k_{1,n}^{m} \
    \label{eqn:precomputed-forward-reduction-step-n} 
\end{align}
%
here $k_{1,2}^m$ and $k_{1,n}^m$ are
precomputed values of $k_1$
for the ``outer'' equations at the $m$th step.

\subsection{Backward substitution}

The backward substitution step for all equations $i > 1$ is

\begin{equation}
x_i = \frac{d^{\prime}_i - a^mx_{i-1} - \
    c^{m}x_{i+1}}{b^m}
\label{eqn:precomputed-backward-substitution-step}
\end{equation}
%
where $a^m$, $b^m$ and $c^m$ are the precomputed
coefficients for $i>1$ at the step $m$.
For the first equation, we have instead:
%
\begin{equation}
x_1 = \frac{d^{\prime}_1 - c^{m}x_{2}}{b_1^m}
\label{eqn:precomputed-backward-substitution-step-1}
\end{equation}
%
where $b_1^m$ is the
coefficient computed for $i = 1$ at step $m$.

\section{Implementation}

\subsection{Precomputing forward reduction coefficients}

\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/cyclic-reduction-precomputing.eps}
\end{center}
\caption{Maximum storage required for forward reduction coefficients
at all steps $m=1, 2, ... log_2(n)-1$}
\label{fig:cyclic-reduction-precomputing}
\end{figure}

Figure \ref{fig:cyclic-reduction-precomputing}
shows the
\emph{maximum} amount of storage required for
storing each of the coefficients
$\{b_1, c_1, a_0, b_0, c_0, a_n, b_n\}$,
or auxiliary variables $k_1$ and $k_2$.
A careful analysis of the
forward reduction and backward substitution equations
reveals that the actual amount of storage is somewhat less.
For instance,
the values $b_n^m$ are unused
in Eqs (\ref{eqn:precomputed-forward-reduction-step})-
(\ref{eqn:precomputed-backward-substitution-step-1}).
The values $a_n^m$ and $b_n^m$ are
required only for solving the $2-by-2$
system at the end of the forward reduction phase,
i.e., for $m=log_2(n)-1$,
and they are not stored for the previous steps.
Similarly, the values $c_1^m$ are equal to $c_0^m$,
and do not require special storage.
The set of precomputed coefficients required to be stored is then
$\{a_0^m, b_0^m, c_0^m, k_1^m, k_2^m, b_1^m, k_{1,1}^m, k_{1,n}^m\}$.
Each of these ``coefficient arrays'' are computed on the CPU
and transferred to the GPU.
Additionally, the two scalars $a_n^{log_2(n)-1}$ and $b_n^{log_2(n)-1}$
are required for the 2-by-2 solve.
%
\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/block-mapping.eps}
\end{center}
\caption{Storing right hand sides and mapping to thread blocks.}
\label{fig:block-mapping}
\end{figure}
%
The right hand sides that the system must be solved for
are stored in a single contiguous array in GPU memory
\ref{fig:block-mapping}.
The precomputed coefficient arrays and the right hand side array
are passed as inputs to the compute kernels that implement
the modified cyclic reduction.
We describe two implementations of the kernels,
the first works entirely with global memory,
and the second leverages the GPU's \emph{shared} memory.

\begin{figure}
\begin{center}
\includegraphics[width=400pt]{img/global-and-shared.eps}
\end{center}
\caption{Thread activity in shared memory (top) and global memory (bottom) implementations.}
\label{fig:global-and-shared}
\end{figure}

\subsection{Global memory implementation}

This implementation works entirely on the GPU's global memory,
Here, we define two kernels - one for the forward reduction step,
and the other for the backward substitution step.
Each kernel is called $log_2(n)-1$ times.
An extra call to the
forward reduction kernel performs
the two-by-two solve.
At each step,
the size of the thread blocks is determined by the \emph{stride}
between elements accessed at that step.
For the forward reduction phase,
we use $n/2$ threads per block for the first step,
$n/4$ threads for the second step, and so on.
The pattern is reversed for the backward substitution phase,
beginning with 2 threads per block for the first step.
Although this ensures that there are no inactive threads
at any stage,
the occupancy of the GPU is still very low
during the end of forward reduction and the end of backward substitution.

\begin{listing}[H]
\begin{minted}[fontsize=\footnotesize]{python}
stride = 1
for i in range(n-1):
    stride *= 2
    forwardReductionKernel(num_blocks=nrhs, block_size=n/stride)

solveTwoByTwo()

for i in range(n-1):
    stride /= 2
    backSubstitutionKernel(num_blocks=nrhs, block=n/stride)
\end{minted}
\end{listing}

The precomputed coefficient arrays and right hand
are accessed by the kernels from global memory.
The kernel suffers from
strided memory access for the right-hand side,
but the precomputed coefficient values are accessed
without major coalescing problems.
Further, by precomputing the forward reduction coefficients,
we greatly reduce the number of computations
(and thus the number of uncoalesced memory accesses).

\subsection{Shared memory implementation}

In the shared memory approach (Fig. \ref{fig:global-and-shared}),
we launch a single kernel to perform the entire
cyclic reduction solve.
The kernel is launched with $n/2$ threads per block.
Each thread block is allocated a block of shared memory of size $n/2$.
Each thread of a block performs the first reduction step
[Eq. (\ref{eqn:precomputed-forward-reduction-step})]
by accessing the required values
$d_i$, $d_{i-1}$ and $d_{i+1}$ from global memory,
storing the result in shared memory.
In subsequent reduction steps,
$d_i$, $d_{i-1}$ and $d_{i+1}$
are accessed from shared memory,
avoiding the uncoalesced global memory accesses
seen in the global memory implementation.
In each back substitution step,
threads overwrite the existing values in shared memory
with the values of the solution.
In the final step,
shared memory is filled completely
with the even-indexed solution values.
Each thread then computes an odd-indexed solution value,
storing it directly in global memory
and copies the even-indexed solution value
from shared memory to global memory.
The entire solution is done in a single kernel launch
to hide the latency of
memory transfers between global and shared memory.
Explicit synchronization between the threads of a block
is required within the kernel at the end of each step.

The shared memory implementation suffers from two major issues:
first, the number of active threads is halved at each forward reduction step,
(and subsequently doubled at each backward reduction step).
Synchronization between threads of a block is necessary at each step.
Thus, a significant portion of the kernel execution time is
spent by idle threads waiting for active threads to complete execution.
Secondly, the strided access to the
right-hand side values leads to bank-conflicts.
However, the number of bank conflicts is significantly smaller
than in cyclic reduction implementations for general tridiagonal systems,
due the reduced number of computations performed.
